Gusto: a toolkit for compatible finite element dynamical cores

### 2024-07-25 source/paper.tex
draft of FML example finished

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of discretisation methods for weather
and climate models, along with enough detail about the finite element
discretiations provided by Gusto to demonstrate the similarities
between them. These similarities guide the design of the code
structure described in the next section and motivate the use of Form
Manipulation Language (FML) that underlies the flexibility of Gusto, a
detailed example of which will be given in section \ref{sec: FML}. The
convention in Gusto is to write the equations in residual form,
meaning that all the terms appear on the left hand side. We will
follow this convention below in contrast to many papers which present
the equations with the pressure gradient term and any forcing terms on
the right hand side.

@@

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value, defaulting to \texttt{True} which indicates the
presence of the label. Labels can also be used to tag parts of the
form. For example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term. The end result of labelling the terms
in the residual is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with an FML
\texttt{label}. Labelled forms can be added together, multiplied by
floats and have additional terms added. More complex manipulations are
enabled by the \texttt{label\_map} method that provides a way of
filtering the terms in the form based on their labels, applying
different mappings to terms that pass or fail the conditions defined
in the filter. A common operation might be to filter for terms that
have a certain label, such as \texttt{transport}, drop terms that do
not have this label and multiply those that do by the timestep. We
shall demonstrate this in the following example.

**describe methods has_label and get

@@

We illustrate the use of FML for this example in three steps: 1)
define and label the weak form of the residual; 2) build the solvers
for the explicit and implicit systems; 3) run the timestepping
loop. Since the intention of this example is to demonstrate FML rather
than the compatible finite element discretisation, we use space of piecewise
quadratic continuous Lagrange polynomials, denoted by $P_2$, for the finite
element discretisation as this greatly simplifies the forms. The
finite element discretisation of equation \ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

Listing \ref{lst:define-residual} presents the code to define this
spatial discretisation of the residual with the $P_2$ function space
denoted by \texttt{V}. Here we omit the code that sets up the mesh,
function spaces and parameters $c$ and $\kappa$, although this can be
found in the supplementary materials. The first two lines define the
test function \texttt{v} and function \texttt{q}, corresponding to $v$
and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
the residual, i.e. the left hand side of equation
\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
\texttt{transport} and \texttt{diffusion} to tag the corresponding
terms. The final line labels the function \texttt{q} as the subject of
this residual, i.e. the prognostic field. Note that the labels used
for tagging just take the UFL form as their argument and their value is
simply the default, \texttt{True}, in contrast to the \texttt{subject}
label that has the value \texttt{q}.

@@

To define the solvers for the explicit and implicit steps we need to create UFL forms for the left and right sides, given by the following equations
\begin{align}
  \label{eq:explicit-advdiff}
  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
  \label{eq:implicit-advdiff}
  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2.
\end{align}
To do this we use the \texttt{label\_map} method to loop over the
terms in the residual making use of the predefined mappings
\texttt{drop} and \texttt{keep} to either remove or retain the term
based on the value of the filter. This code is given in listings
\ref{lst:define-explicit} and \ref{lst:define-implicit}. In listing
\ref{lst:define-explicit} we create the left and right hand sides of
equation \ref{eq:explicit-advdiff}. The left hand side should only
contain the time derivative term so the filter checks if each term has
the label \texttt{time\_derivative}, keeping the one that does and
dropping all others. To make this the left hand side of a
\texttt{LinearVariationalProblem}, we replace the subject with a trial
function using the predefined \texttt{replace\_subject} function, a
lightweight wrapper around the UFL replace method which extracts the
subject of the term and replaces it with the new expression, in this
case simply a trial function \texttt{q\_trial}. \JScomment{Is the
  labelling for the rhs the clearest? maybe we should keep the terms
  that we want rather than assuming that dropping diffusion gives the
  right thing?}

@@

@@

\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}

>>>>

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of discretisation methods for weather
and climate models, along with enough detail about the finite element
discretiations provided by Gusto to demonstrate the similarities
between them. These similarities guide the design of the code
structure described in the next section and motivate the use of Form
Manipulation Language (FML) that underlies the flexibility of Gusto, a
detailed example of which will be given in section \ref{sec: FML}. The
convention in Gusto is to write the equations in residual form,
meaning that all the terms appear on the left side. We will follow
this convention below in contrast to many papers which present the
equations with the pressure gradient term and any forcing terms on the
right side.

@@

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value, defaulting to \texttt{True} to indicate the presence
of the label. Labels can also be used to tag parts of the form. For
example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term. The end result of labelling the terms
in the residual is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with FML
labels. Labelled forms can be added together, multiplied by floats and
have additional terms added. More complex manipulations are enabled by
the \texttt{label\_map} method that provides a way of filtering the
terms in the form based on their labels, applying different mappings
to terms that pass or fail the conditions defined in the filter. A
common operation might be to filter for terms that have a certain
label, such as \texttt{transport}, drop terms that do not have this
label and multiply those that do by the timestep. We shall demonstrate
this in the following example.

\JScomment{describe methods has label and get?}

@@

We illustrate the use of FML for this example in three steps: 1)
define and label the weak form of the residual; 2) build solvers for
the explicit and implicit systems; 3) run the timestepping loop. Since
the intention of this example is to demonstrate FML rather than the
compatible finite element discretisation, we use space of piecewise
quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
finite element discretisation as this greatly simplifies the
forms. The finite element discretisation of equation
\ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

Listing \ref{lst:define-residual} presents the code to define this
spatial discretisation of the residual with the $P_2$ function space
denoted by \texttt{V}. Here we omit the code that sets up the mesh,
function spaces and parameters $c$ and $\kappa$, although this can be
found in the supplementary materials. The first two lines define the
test function \texttt{v} and function \texttt{q}, corresponding to $v$
and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
the residual, i.e. the left side of equation \ref{eq:weak-advdiff},
using the FML labels \texttt{time\_derivative}, \texttt{transport} and
\texttt{diffusion} to tag the corresponding terms. The final line
labels the function \texttt{q} as the subject of this residual,
i.e. the prognostic field. Note that the labels used for tagging just
take the UFL form as their argument and their value is simply the
default, \texttt{True}, in contrast to the \texttt{subject} label that
has the value \texttt{q}.

@@

To define the solvers for the explicit and implicit steps we need to create UFL forms for the left and right sides, given by the following equations
\begin{align}
  \label{eq:explicit-advdiff}
  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
  \label{eq:implicit-advdiff}
  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2.
\end{align}
To do this we use the \texttt{label\_map} method to loop over the
terms in the residual making use of the predefined mappings
\texttt{drop} and \texttt{keep} to either remove or retain the term
based on the value of the filter. This code is given in listings
\ref{lst:define-explicit} and \ref{lst:define-implicit}. In listing
\ref{lst:define-explicit} we create the left and right sides of
equation \ref{eq:explicit-advdiff}. The left side should only contain
the time derivative term so the filter checks if each term has the
label \texttt{time\_derivative}, keeping the one that does and
dropping all others. To make this the left side of a
\texttt{LinearVariationalProblem}, we replace the subject with a trial
function using the predefined \texttt{replace\_subject} function, a
lightweight wrapper around the UFL replace method which extracts the
subject of the term and replaces it with the new expression, in this
case simply a trial function \texttt{q\_trial}. The right side of the
explicit solver has two forms, one coming from the discretisation of
the time derivative and the other from the transport term. Both are
evaluated using the values of $q$ at time step $n$. We build this by
first dropping the diffusion term (line 6) and then multiplying the
terms that do not have the \texttt{time\_derivative} label by $-\Delta
t$ (line 8), where the minus sign is required to move the term from
the left side to the right side. The final step is to replace the
subject with the function \texttt{qn} that holds the values of $q$ at
time step $n$. We can then extract the UFL forms from
\texttt{lhs\_explicit} and \texttt{lhs\_implicit} to create a
\texttt{LinearVariationalProblem} and corresponding
\texttt{LinearVariationalSolver} that, when solved, will put the
result into the function \texttt{qstar}.

\JScomment{Is the
  labelling for the rhs the clearest? maybe we should keep the terms
  that we want rather than assuming that dropping diffusion gives the
  right thing?}

@@

The implicit solver is defined in a similar way (see listing
\ref{lst:define-implicit}). To create the left side of equation
\ref{eq:implicit-advdiff} we begin with the original residual, drop
the transport term (line 1), multiply all remaining terms that are not
the time derivative by $\Delta t$ (line 3) and replace the subject
with a trail function (line 5). The right side only contains the form
coming from the discretisation of the time derivative (line 7),
evaluated using the result of the explicit solve, \texttt{qstar} (line
9).

@@

The final part of the code is the time loop which calls the solve
method of the explicit and then the impicit solvers defined above, see
listing \ref{lst:timeloop}

\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}
